Choosing Color Palettes for Statistical Graphics

### 2007-01-04 trunk/vcd/paper/Color/TAS.Rnw
first draft for TAS revision finished
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@635 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass{Z}
%% need no \usepackage{Sweave}
\SweaveOpts{engine=R, eps=FALSE, echo=FALSE, results=hide}
\usepackage{amsfonts}

@@

To show what can be gained by selecting appropriate color schemes, we present a collection
of illustrations with typical color graphics based on HSV palettes (always in the left panels)
and more suitable HCL palettes (always in the right panels). All HSV color palettes (and some
of the data sets) are taken from recent publications in statistical journals such as
\textit{Journal of the American Statistical Association},
\textit{Journal of Computational and Graphical Statistics} or
\textit{Computational Statistics \& Data Analysis}. The examples have been selected to provide
an overview of (typically colored) statistical graphics and their respective pitfalls (when
it comes to color palette choice). All graphics are in principle very useful but can be even enhanced
by using better color schemes.

Our first illustration is a heatmap, a very popular display for visualizing a function of
two arguments. Here, we use a bivariate kernel density estimate \citep[for the Old
Faithful geyser eruptions data from][]{vcd:Azzalini+Bowman:1990}---other typical applications
include objective functions with two arguments \citep[as in][]{vcd:Gneiting+Sevcikova+Percival:2006}
or (physical) measurements on a 2-dimensional grid \citep[as in][]{vcd:Yang+Buckley+Dudoit:2002}.
Figure~\ref{fig:bkde} shows such a heatmap bringing out the relationship between the duration
of an eruption of the Old Faithful geyser in Yellowstone National Park 
and the waiting time for this eruption. It reveals a multi-modal distribution: short waiting
times (around 50 minutes) are typically followed by a long eruption (around 4 minutes)
whereas long waiting times (around 80 minutes) can be followed by either a long
or short eruption (around 4 minutes).

@@

A simple and very effective palette for such a display is a set of gray colors
as in the top right panel of Figure~\ref{fig:bkde}. This is often (appropriately) used
in printed papers when the journal does not offer color graphics---however, in journals
that support color graphics, on presentation slides and in interactive usage (in statistical
software packages), many users prefer to have colored displays and most often HSV
palettes (as in the left two panels) are used. These palettes code the variable of interest
by varying hue in an HSV color wheel as done by \citet[Figure~4a]{vcd:Yang+Buckley+Dudoit:2002}
or \citet[Figures~1--4]{vcd:Gneiting+Sevcikova+Percival:2006}. The palette in the upper left panel
codes increasing density by going from a blue to a red hue (via green and yellow)---a similar
strategy are the ``heat colors'' in the lower left panel that increase from yellow to red.
The latter works a bit better than the former, however both palettes exhibit several drawbacks.
The modes in the map look much more like ``rings'' than a smoothly increasing/decreasing 
density. The heatmap looks very flashy which---although good for drawing attention to a plot---makes
it hard to hold the attention of the viewer for a longer time because the large areas shaded
with saturated colors can be distracting and produce after-image effects \citep{vcd:Ihaka:2003}.

@@

To illustrate sequential palettes,

 Both versions of the heatmap use a sequential
palette as derived in Section~\ref{sec:sequential} balanced towards the same gray levels
with $L \in [30, 90]$ and $p_2 = 1.5$. The sequential palette in the left panel uses only gray
colors (i.e., $C_{\max} = 0$) and the palette in the right panel additionally employs
colors with $H \in [0, 90]$, $C \in [30, 80]$ and $p_1 = 1/5$. The parameters have been 
chosen such that the peaks in the density are brought out very clearly by using a rather
large luminance contrast and $p_2 > 1$.

@@

Figure~\ref{fig:nigeria} is taken from \citep[Figure~5, left]{vcd:Kneib:2006}

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<seats, fig=TRUE, height=4, width=11>>=
par(mar = rep(0.8, 4), mfrow = c(1, 2))
seats <- structure(c(226, 61, 54, 51, 222),
  .Names = c("CDU/CSU", "FDP",  "Linke", "Gruene", "SPD"))
parties_hcl <- rainbow_hcl(6, c = 60, l = 75)[c(5, 2, 6, 3, 1)]
parties_hsv <- hsv(c(0, 60, 300, 120, 0)/360, v = c(0, 1, 1, 1, 1))
names(parties_hcl) <- names(parties_hsv) <- names(seats)
names(seats)[4] <- "Grüne"
pie(seats, clockwise = TRUE, col = parties_hsv, radius = 1)
pie(seats, clockwise = TRUE, col = parties_hcl, radius = 1)
@
\caption{\label{fig:seats} Seats in the German parliament.}
\end{center}
\end{figure}

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<votes, fig=TRUE, height=6, width=14>>=
data("Bundestag2005")
votes <- Bundestag2005[c(1, 3:5, 9, 11, 13:16, 2, 6:8, 10, 12),
                       c("CDU/CSU", "FDP", "SPD", "Gruene", "Linke")]
parties_hcl <- parties_hcl[colnames(votes)]
parties_hsv <- parties_hsv[colnames(votes)]
dimnames(votes)[[2]][c(4, 5)] <- c("Gr", "Li")

grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 2)))
pushViewport(viewport(layout.pos.col=1, layout.pos.row=1))
mosaic(votes, gp = gpar(fill = parties_hsv),
  spacing = spacing_highlighting, labeling = labeling_left,
  labeling_args = list(rot_labels = c(0, 90, 0, 0), pos_labels = "center",
  just_labels = c("center","center","center","right"), varnames = FALSE),
  margins = unit(c(2.5, 1, 1, 12), "lines"),
  keep_aspect_ratio = FALSE, newpage = FALSE)
popViewport()
pushViewport(viewport(layout.pos.col=2, layout.pos.row=1))
mosaic(votes, gp = gpar(fill = parties_hcl),
  spacing = spacing_highlighting, labeling = labeling_left,
  labeling_args = list(rot_labels = c(0, 90, 0, 0), pos_labels = "center",
  just_labels = c("center","center","center","right"), varnames = FALSE),
  margins = unit(c(2.5, 1, 1, 12), "lines"),
  keep_aspect_ratio = FALSE, newpage = FALSE)
popViewport(2)
@
\caption{\label{fig:votes} Votes in the German election 2005 (by province).}
\end{center}
\end{figure}

The first example illustrates qualitative palettes
and visualizes data from the 2005 election for the German parliament ``Bundestag''.
In this election, five parties were able to obtain enough votes to enter the
Bundestag---the distribution of seats is depicted in a pie chart in Figure~\ref{fig:seats}.
The colors used are rough metaphors for the political parties, using a red hue $H = 0$ for the
social democrats SPD, a blue hue $H = 240$ for the conservative CDU/CSU, a yellow hue $H = 60$
for the liberal FDP, a green hue $H = 120$ for the green party ``Die Gr\"unen'' and a purple 
hue $H = 300$ for the leftist party ``Die Linke''. To obtain rather intense colors, we set
chroma to $C = 60$ and luminance to $L = 75$ for all hues
leading to a much better balanced palette compared to the fully saturated
colors usually found in the (German) media (which instead of blue typically use the extreme
``color'' black for the CDU/CSU). The pie chart clearly shows that neither the governing
coalition of SPD and Gr\"une nor the opposition of CDU/CSU and FDP could assemble a majority.
Given that no party would enter a coalition with the leftists, this lead to a big coalition
of CDU/CSU and SPD. Figure~\ref{fig:votes} shows the distribution of votes in this election
stratified by province (Bundesland) in a highlighted mosaic display. The order of provinces is from 
north to south, first for the 10 western provinces (the former Federal Republic of Germany, FRG),
then for the 6 eastern provinces (the former German Democratic Republic, GDR). 
Clearly, the SPD performed better in the north and the CDU/CSU better in the south; furthermore,
Die Linke performed particularly well in the eastern provinces and in Saarland.

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<scatter>>=
if(generate_png) {
  set.seed(123)
  vc <- function(var1, var2, cov)
    matrix(c(var1, cov * sqrt(var1 * var2), cov * sqrt(var1 * var2), var2), ncol = 2)
  dat <- rbind(rmvnorm(1000, c(0, 0), vc(0.05, 0.3, -0.95)),
               rmvnorm(1000, c(1.8, 2.5), vc(0.12, 0.25, 0)),
               rmvnorm(1000, c(3.8, 3.5), vc(0.02, 0.15, -0.8)))
  dat <- data.frame(y = dat[,2], x = dat[,1], id = rep(1:3, rep(1000, 3)))

@@

similar to \citet[Figure~3]{vcd:Celeux+Hurn+Robert:2000}

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<haireye, fig=TRUE, height=5.5, width=12>>=
he <- margin.table(HairEyeColor, 1:2)
grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 2)))
pushViewport(viewport(layout.pos.col=1, layout.pos.row=1))
mosaic(he, gp = shading_hsv, split_vertical = TRUE, newpage = FALSE, legend = FALSE)
popViewport()
pushViewport(viewport(layout.pos.col=2, layout.pos.row=1))
mosaic(he, gp = shading_hcl, split_vertical = TRUE, newpage = FALSE, legend = FALSE)
popViewport(2)
@
\caption{\label{fig:haireye} Extended mosaic display for haireye data.}
\end{center}
\end{figure}

taken from \citet[Figure~11, left/middle]{vcd:Friendly:2002}

As pointed out in Section~\ref{sec:diverging}, diverging palettes are particularly useful
when visualizing residuals or correlations (with natural neutral value 0) or probabilities 
in 2-class supervised learning (with neutral value 0.5). Examples for both situations are
provided here. Figure~\ref{fig:haireye} visualizes the outcome of a double-blind clinical
trial investigating a new treatment for rheumatoid arthritis. The mosaic rectangles alone
signal that the treatment lead to higher improvement compared to the placebo group; this effect
is shown to be significant by the shading that codes the size of the Pearon residuals.
Positive residuals, corresponding
to more observations in the corresponding cell than expected under independence, are depicted
in blue, negative residuals in red. Light colors signal significance at 10\% level, full colors
significance at 1\% level. Hence, it can be concluded that there are significantly more
marked improvements in the treated group and significantly fewer in the placebo group than
would be expected under independence between treatment and improvement. More details can
be found in \cite{vcd:Zeileis+Meyer+Hornik:2007}.

Both diverging palettes employ a smaller contrast on the luminance axis ($L \in [50, 90]$)
and a larger chroma contrast ($C \in [0, 100]$) compared to the sequential palettes from
the previous example. 
For Figure~\ref{fig:haireye}, the choice of
$p_1$ and $p_2$ is less important because only 5 colors are selected---we use values
$p_1 = p_2 = 1$.

@@

A popular implementation of such a color space, available in many graphics and
statistics software packages, are HSV colors. They are a simple transformation of RGB colors
and are defined by a triplet $(H, S, V)$ with $H \in [0, 360]$ and $S, V \in [0, 100]$.
%% HSV space has the shape of a single regular cone (often inflated to a regular
%% cylinder). Vertical sections through this space are shown in the upper panel
%% of Figure~\ref{fig:hsv-hcl}, depicting hue and saturation given different value levels.
Although simple to specify and easily available in many computing environments,
HSV colors have a fundamental drawback: its three dimensions map to the three
dimensions of human color perception very poorly. The three dimensions are
confounded:
%% which is most easily seen when converting the vertical sections to 
%% gray scale images in Figure~\ref{fig:hsv-hcl-gray}. Clearly,
the brightness of
colors is not uniform over hues and saturations (given value)---therefore,
HSV colors are often not considered to be perceptually based.

@@

\cite{vcd:Ihaka:2003} describes a simple strategy for choosing such palettes: chroma
and luminance are kept fixed and only the hue is varied for obtaining different colors
which are consequently all balanced towards the same gray. As Figure~\ref{fig:hsv-hcl}
illustrates, the range of hues available depends on the combination of chroma and
luminance chosen. Figure~\ref{fig:pal-q-construct} depicts how three colors
are chosen, given $C = 50$ and $L = 70$.

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<wheel, fig=TRUE, height=4, width=11>>=
par(mar = rep(0.8, 4), mfrow = c(1, 2))
par(mfrow = c(1,2), mar = c(1,1,1,1), oma = c(0,0,0,0))
pie(rep(1,6), radius = 1, col = rainbow(6), labels = 360 * 0:5/6)
pie(rep(1,6), radius = 1, col = rainbow_hcl(6, c = 50, l = 70), labels = 360 * 0:5/6)
@
\caption{\label{fig:wheel} HSV-based and HCL-based color wheel.}
\end{center}
\end{figure}

Various strategies for choosing the hues in a certain palette are conceivable.
A simple and intuitive one is to use colors as metaphors for categories
(e.g., for political parties), another approach would be to use segments from
the color wheel corresponding to nearby or distant colors. The latter is shown
in Figure~\ref{fig:pal-q-examples} which depicts
examples for generating qualitative sets of colors $(H, 50, 70)$.
In the upper left panel colors from the full spectrum are used ($H = 30, 120, 210, 300$)
creating a `dynamic' set of colors. The upper right panel shows a `harmonic'
set with $H = 60, 120, 180, 240$. Warm colors (from the blue/green part of the spectrum: 
$H = 270, 230, 190, 150$) and cold colors (from the yellow/red part of the spectrum:
$H = 90, 50, 10, 330$) are shown in the lower left and right panel, respectively.

@@

@@

The simplest solution to this task is to employ light/dark contrasts, i.e., rely
on the most basic perceptual axis. The interestingness can be coded
by an increasing amount of gray corresponding to decreasing luminance in HCL space:
  \[ (H, 0, 90 - i \cdot 60), \]
where the hue $H$ used does not matter, chroma is set to 0 (i.e., no color), and 
luminance ranges in $[30, 90]$ avoiding the extreme colors white ($L = 100$) and 
black ($L = 0$). Instead of going linearly from light to dark gray, luminance
could also be increased nonlinearly, e.g., by some function $f(i)$ that
controls whether luminance is increased quickly with intensity or not. We found $f(i) = i^p$ 
to be a convenient transformation where the power $p$ can be varied to achieve
different degrees of non-linearity.

Furthermore, the intensity $i$ could additionally be coded by colorfulness (chroma), e.g.,
  \[ (H, 0 + i^p \cdot C_{\max}, L_{\max} - i^p \cdot (L_{\max} - L_{\min})). \]
This strategy is depicted in the left panel of Figure~\ref{fig:pal-s-construct} for a blue
hue $H = 260$ and different combinations of maximal chroma ($C_{\max} = 0$, $80$ and $100$,
respecitvely) and minimal luminance ($L_{\min} = 30$, $30$ and $50$, respectively).
The first two combinations are also shown in the first two rows of Figure~\ref{fig:pal-s-examples}.
The right panel of Figure~\ref{fig:pal-s-construct} shows that the exact same strategy
is not possible  for the green hue $H = 120$. While the gray colors without chroma can be chosen in 
the same way, there is a stronger trade-off between using dark colors (with low luminance)
and colorful colors (with high chroma). Hence, the second path from light gray to 
full green ends at a much lighter color with $L_{\min} = 75$.

@@

To increase the contrast between the colors in the palette even further, the ideas from
the previous sequential palettes can also be combined with qualitative palettes by
simultaneously varying the hue as well:
  \[ (H_2 - i \cdot (H_1 - H_2), C_{\max} - i^{p_1} \cdot (C_{\max} - C_{\min}),
      L_{\max} - i^{p_2} \cdot (L_{\max} - L_{\min})). \]
One application would be an HCL-based version of ``heat colors'' that increase from a light yellow
(e.g., $(90, 30, 90)$) to a full red (e.g., $(0, 100, 50)$). To make the change in hue
visible, the chroma needs to increase rather quickly for low values of $i$ and then only
slowly for higher values of $i$. This can be achieved by choosing a power $p_1 < 1$.
Such a strategy is adopted for the palettes shown in the lower three rows in
Figure~\ref{fig:pal-s-examples} using different pairs of hues as well as different
chroma and luminance contrasts. The exact parameter combinations are provided in the
appendix.

@@

Diverging palettes are also used for coding numerical information ranging in a certain
interval---however, this interval includes a neutral value. Examples for this include
residuals or correlations (both with the neutral value 0) or binary classification
probabilities (with neutral value 0.5) that could be visualized in mosaic plots (see Figure~\ref{fig:haireye})
or classification maps (see Figure~\ref{fig:class}). Analogously to the previous
section, we suppose that we want to visualize an intensity or interestingness $i$
from the interval $[-1, 1]$ (without loss of generality).

@@

%%FIXME: comment on vcd:Lumley:2006

@@

Many statistical graphics---especially when displayed on a computer screen,
e.g., as in interactive usage, electronic papers or presentation slides---
employ colors to code information about a certain variable. Despite this
omnipresence of color, there is often only little guidance in statistical
software packages on how to choose a palette appropriate for a particular
visualization task---auspicious tools such as \pkg{ColorBrewer.org} notwithstanding.
We try to address this problem by suggesting color schemes for coding
categorical information (qualitative palettes) and numerical information
(sequential and diverging palettes) based on the perceptually-based
HCL color space.

>>>>

\documentclass{Z}
%% need no \usepackage{Sweave}
\SweaveOpts{engine=R, eps=FALSE, echo=FALSE, results=hide}
\usepackage{amsfonts,rotating}

@@

To show what can be gained by selecting appropriate color schemes, we present a collection
of illustrations with typical color graphics based on HSV palettes (always in the left panels)
and more suitable HCL palettes (always in the right panels). All HSV color palettes (and some
of the data sets) are taken from recent publications in statistical journals such as
\textit{Journal of the American Statistical Association},
\textit{Journal of Computational and Graphical Statistics} or
\textit{Computational Statistics \& Data Analysis}. The examples have been selected to provide
an overview of typically colored statistical graphics and their respective pitfalls (concerning
color palette choice).
%% All graphics are in principle very useful but can be even enhanced
%% by using better color schemes.

Our first illustration is a heatmap, a very popular display for visualizing a scalar function of
two arguments. Here, we use a bivariate kernel density estimate \citep[for the Old
Faithful geyser eruptions data from][]{vcd:Azzalini+Bowman:1990}---other typical applications
include objective functions with two arguments \citep[as in][]{vcd:Gneiting+Sevcikova+Percival:2006}
or (physical) measurements on a 2-dimensional grid \citep[as in][]{vcd:Yang+Buckley+Dudoit:2002}.
Figure~\ref{fig:bkde} shows such a heatmap bringing out the relationship between the duration
of an eruption of the Old Faithful geyser in Yellowstone National Park 
and the waiting time for this eruption. It reveals a multi-modal distribution: short waiting
times (around 50 minutes) are typically followed by a long eruption (around 4 minutes)
whereas long waiting times (around 80 minutes) can be followed by either a long
or short eruption (around 4 minutes).

@@

A simple and very effective palette for such a display is a set of gray colors
as in the top right panel of Figure~\ref{fig:bkde}. This is often (appropriately) used
in printed papers when the journal does not offer color graphics---however, in journals
that support color graphics (or on presentation slides and in interactive usage in statistical
software packages), many users prefer to have colored displays and most often use HSV
palettes (as in the two left two panels). These palettes code the variable of interest
by varying hue in an HSV color wheel as done by \citet[Figure~4a]{vcd:Yang+Buckley+Dudoit:2002}
or \citet[Figures~1--4]{vcd:Gneiting+Sevcikova+Percival:2006}. The palette in the upper left panel
codes increasing density by going from a blue to a red hue (via green and yellow)---a similar
strategy are the ``heat colors'' in the lower left panel that increase from yellow to red.
The latter works a bit better than the former, however both palettes exhibit several drawbacks.
The modes in the map look much more like ``rings'' rather than a smoothly increasing/decreasing 
density. The heatmap looks very flashy which---although good for drawing attention to a plot---makes
it hard to hold the attention of the viewer for a longer time because the large areas shaded
with saturated colors can be distracting and produce after-image effects \citep{vcd:Ihaka:2003}.

@@

@@

A similar example is presented in Figure~\ref{fig:nigeria}
\citep[taken from][Figure~5, left]{vcd:Kneib:2006}, depicting posterior mode estimates
for childhood mortality in different regions of Nigeria. Deviations from a model for
childhoold mortality are brought out by shading a map according to the corresponding
residuals, revealing decreased mortality in the south-west and increased mortality in
the north-east. \cite{vcd:Kneib:2006} uses an HSV-based palette coding the residuals by
the hue, going from green via yellow to red. Our HCL-based palette also employs green
and red hues for negative and postive residuals respectively, but codes neutral values
(around 0) by a neutral light gray. Compared to the HSV-based palette, this offers again
a number of advantages: only the interesting areas are highlighted by high-chroma colors;
flashy colors are avoided making it easier to look at the display for a longer time; positive
and negative residuals with the same absolute size receive the same perceptual weight
by being balanced towards the same gray.

\setkeys{Gin}{width=\textwidth}
\begin{figure}[t!]
\begin{center}
<<seats, fig=TRUE, height=4, width=11>>=
par(mar = rep(0.8, 4), mfrow = c(1, 2))
seats <- structure(c(226, 61, 54, 51, 222),
  .Names = c("CDU/CSU", "FDP",  "Linke", "Gruene", "SPD"))
parties_hcl <- rainbow_hcl(6, c = 60, l = 75)[c(5, 2, 6, 3, 1)]
parties_hsv <- hsv(c(0, 60, 300, 120, 0)/360, v = c(0, 1, 1, 1, 1))
names(parties_hcl) <- names(parties_hsv) <- names(seats)
names(seats)[4] <- "Grüne"
pie(seats, clockwise = TRUE, col = parties_hsv, radius = 1)
pie(seats, clockwise = TRUE, col = parties_hcl, radius = 1)
@
\caption{\label{fig:seats} Seats in the German parliament.}
\end{center}
\end{figure}

Although problematic for many tasks, pie charts can be useful for visualizing
whether a set of pie segments constitutes a majority. A typical application is shown
in Figure~\ref{fig:seats}, visualizing the distribution of seats in the German parliament
``Bundestag'' following the 2005 election. In this election, five parties were able to
obtain enough votes to enter the Bundestag---however, neither the governing
coalition of SPD and Gr\"une nor the opposition of CDU/CSU and FDP could assemble a majority.
Given that no party would enter a coalition with the leftists ``Die Linke'', this lead to
a big coalition of CDU/CSU and SPD. In graphical displays, the parties are usually matched
by using colors as metaphors: red for the social democrats SPD, black for the conservative
CDU/CSU, yellow for the liberal FDP, green for the green party ``Die Gr\"unen'' and purple 
for the leftist party ``Die Linke''. The left panel shows fully saturated HSV colors as usually
found in the (German) media whereas the right panel uses less flashy HCL colors with the
same hues (except for the CDU/CSU where a blue hue instead of the extreme ``color'' black
is used). The advantage of the latter is again that they are easier to hold in focus for
a longer time. Furthermore, they are all balanced towards the same gray and have the same
amount of color, resulting in a perceptually balanced palette that does not introduce undesired
graphical distractions. While it could be argued that this pie chart is such a simple
display that a well-balanced palette is not so important and that flashy colors are to
be preferred, the balancing properties of the HCL-based palette are very important in more
complex displays such as the mosaic display in Figure~\ref{fig:votes}. This re-uses the same
colors and shows the distribution of votes in the 2005 election
stratified by province (Bundesland). The order of provinces is from 
north to south, first for the 10 western provinces (the former Federal Republic of Germany, FRG),
then for the 6 eastern provinces (the former German Democratic Republic, GDR). 
Clearly, the SPD performed better in the north and the CDU/CSU better in the south; furthermore,
Die Linke performed particularly well in the eastern provinces and in Saarland.

\setkeys{Gin}{width=\textwidth}
\begin{sidewaysfigure}[p]
\begin{center}
<<votes, fig=TRUE, height=6, width=14>>=
data("Bundestag2005", package = "vcd")
votes <- Bundestag2005[c(1, 3:5, 9, 11, 13:16, 2, 6:8, 10, 12),
                       c("CDU/CSU", "FDP", "SPD", "Gruene", "Linke")]
parties_hcl <- parties_hcl[colnames(votes)]
parties_hsv <- parties_hsv[colnames(votes)]
dimnames(votes)[[1]][c(9, 16)] <- c("Baden-Württemberg", "Thüringen")
dimnames(votes)[[2]][c(4, 5)] <- c("Gr", "Li")

grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 2)))
pushViewport(viewport(layout.pos.col=1, layout.pos.row=1))
mosaic(votes, gp = gpar(fill = parties_hsv),
  spacing = spacing_highlighting, labeling = labeling_left,
  labeling_args = list(rot_labels = c(0, 90, 0, 0), pos_labels = "center",
  just_labels = c("center","center","center","right"), varnames = FALSE),
  margins = unit(c(2.5, 1, 1, 12), "lines"),
  keep_aspect_ratio = FALSE, newpage = FALSE)
popViewport()
pushViewport(viewport(layout.pos.col=2, layout.pos.row=1))
mosaic(votes, gp = gpar(fill = parties_hcl),
  spacing = spacing_highlighting, labeling = labeling_left,
  labeling_args = list(rot_labels = c(0, 90, 0, 0), pos_labels = "center",
  just_labels = c("center","center","center","right"), varnames = FALSE),
  margins = unit(c(2.5, 1, 1, 12), "lines"),
  keep_aspect_ratio = FALSE, newpage = FALSE)
popViewport(2)
@
\caption{\label{fig:votes} Votes in the German election 2005 (by province).}
\end{center}
\end{sidewaysfigure}

Color choice is usually much more important in graphical displays with
shaded areas compared to displays with only points or lines. However, color choice
becomes much more relevant also in scatter plots when there are many points. As an
example, Figure~\ref{fig:scatter} depicts a scatter plot with three clusters similar to 
the one shown in \citet[Figure~3]{vcd:Celeux+Hurn+Robert:2000}. To indicate cluster
membership, three different colors (green, yellow, red) are employed. The HSV colors in
the left panel are again very flashy and differ strongly with respect to luminance: the
yellow is much lighter and hardly visible. In contrast, the HCL colors in the right
panel use the same hues, but are balanced with respect to chroma and luminance, i.e.,
the amount of color and gray.

As our final example, Figure~\ref{fig:haireye} \citep[taken from][Figure~11, left/middle]{vcd:Friendly:2002}
visualizes the cross-tabulation of hair end eye color of 592 students in a mosaic display.
Clearly, hair and eye color are not independent and the pattern of association is highlighted
by means of residual-based shading. Cells associated with Pearson residuals whose absolute value exceeds
(2 or) 4 are shaded (light) blue/red. This shows that there are significantly more students
with black hair and brown eyes, blond hair and blue eyes, red hair and green eyes than expected
under independence. Conversely, fewer students than expected have blond hair and brown/hazel eyes
or black hair and blue eyes. Comparing the HSV and HCL colors in Figure~\ref{fig:haireye},
it is shown again that the HCL colors are better balanced (between the red and blue colors)
and less flashy. Even the cells associated with small residuals (below 2) are a bit easier
to read when shaded in a light grey rather than white. More details and extensions to residual-based
shadings can be found in \cite{vcd:Zeileis+Meyer+Hornik:2007}.

\setkeys{Gin}{width=\textwidth}
\begin{figure}[h!]
\begin{center}
<<scatter>>=
if(generate_png) {
  set.seed(123)
  vc <- function(var1, var2, cov)
    matrix(c(var1, cov * sqrt(var1 * var2), cov * sqrt(var1 * var2), var2), ncol = 2)
  dat <- rbind(rmvnorm(1000, c(0, 0), vc(0.05, 0.3, -0.95)),
               rmvnorm(1000, c(1.8, 2.5), vc(0.12, 0.25, 0)),
               rmvnorm(1000, c(3.8, 3.5), vc(0.02, 0.15, -0.8)))
  dat <- data.frame(y = dat[,2], x = dat[,1], id = rep(1:3, rep(1000, 3)))

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[h!]
\begin{center}
<<haireye, fig=TRUE, height=5.5, width=12>>=
he <- margin.table(HairEyeColor, 1:2)
grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 2)))
pushViewport(viewport(layout.pos.col=1, layout.pos.row=1))
mosaic(he, gp = shading_hsv, split_vertical = TRUE, newpage = FALSE, legend = FALSE)
popViewport()
pushViewport(viewport(layout.pos.col=2, layout.pos.row=1))
mosaic(he, gp = shading_hcl, split_vertical = TRUE, newpage = FALSE, legend = FALSE)
popViewport(2)
@
\caption{\label{fig:haireye} Extended mosaic display for haireye data.}
\end{center}
\end{figure}

@@

A popular implementation of such a color space, available in many graphics and
statistics software packages, are HSV colors. They are a simple transformation of RGB colors
and are defined by a triplet $(H, S, V)$ with $H \in [0, 360]$ and $S, V \in [0, 100]$.
%% HSV space has the shape of a single regular cone (often inflated to a regular
%% cylinder). Vertical sections through this space are shown in the upper panel
%% of Figure~\ref{fig:hsv-hcl}, depicting hue and saturation given different value levels.
Although simple to specify and easily available in many computing environments,
HSV colors have a fundamental drawback: its three dimensions map to the three
dimensions of human color perception very poorly. The three dimensions are
confounded:
%% which is most easily seen when converting the vertical sections to 
%% gray scale images in Figure~\ref{fig:hsv-hcl-gray}. Clearly,
The brightness of
colors is not uniform over hues and saturations (given value)---therefore,
HSV colors are often not considered to be perceptually based.

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[b!]
\begin{center}
<<wheel, fig=TRUE, height=4, width=11>>=
par(mar = rep(0.8, 4), mfrow = c(1, 2))
par(mfrow = c(1,2), mar = c(1,1,1,1), oma = c(0,0,0,0))
pie(rep(1,6), radius = 1, col = rainbow(6), labels = 360 * 0:5/6)
pie(rep(1,6), radius = 1, col = rainbow_hcl(6, c = 50, l = 70), labels = 360 * 0:5/6)
@
\caption{\label{fig:wheel} HSV-based and HCL-based color wheel.}
\end{center}
\end{figure}

@@

\cite{vcd:Ihaka:2003} describes a simple strategy for choosing such palettes: chroma
and luminance are kept fixed and only the hue is varied for obtaining different colors
which are consequently all balanced towards the same gray. This effect is illustrated
in Figure~\ref{fig:wheel} which shows a color wheel obtained by varying the hue only in
HSV coordinates $(H, 100, 100)$ and HCL coordinates $(H, 50, 70)$. Clearly, not only the
hue but also the amount of chroma and luminance varies for the HSV wheel.

Figure~\ref{fig:pal-q-construct} depicts how the HCL-based color wheel is constructed.
It shows the hue/chroma plane of HCL space given a luminance of $L = 70$. Not all
combinations of hue and chroma are admissible, however for a chroma of $C = 50$ a full
color wheel can be obtained.
For choosing the hues in a certain palette, various strategies are conceivable.
A simple and intuitive one is to use colors as metaphors for categories
(e.g., for political parties as in Figures~\ref{fig:seats} and \ref{fig:votes}),
another approach would be to use segments from
the color wheel corresponding to nearby or distant colors. The latter is shown
in Figure~\ref{fig:pal-q-examples} which depicts
examples for generating qualitative sets of colors $(H, 50, 70)$.
In the upper left panel colors from the full spectrum are used ($H = 30, 120, 210, 300$)
creating a `dynamic' set of colors. The upper right panel shows a `harmonic'
set with $H = 60, 120, 180, 240$. Warm colors (from the blue/green part of the spectrum: 
$H = 270, 230, 190, 150$) and cold colors (from the yellow/red part of the spectrum:
$H = 90, 50, 10, 330$) are shown in the lower left and right panel, respectively.

@@

The simplest solution to this task is to employ light/dark contrasts, i.e., rely
on the most basic perceptual axis. The interestingness can be coded
by an increasing amount of gray corresponding to decreasing luminance in HCL space:
  \[ (H, 0, 90 - i \cdot 60), \]
where the hue $H$ used does not matter, chroma is set to 0 (i.e., no color), and 
luminance ranges in $[30, 90]$ avoiding the extreme colors white ($L = 100$) and 
black ($L = 0$). Instead of going linearly from light to dark gray, luminance
could also be increased nonlinearly, e.g., by some function $i' = f(i)$ that
controls whether luminance is increased quickly with intensity or not. We found $i' = f(i) = i^p$ 
to be a convenient transformation where the power $p$ can be varied to achieve
different degrees of non-linearity.

Furthermore, the intensity $i$ could additionally be coded by colorfulness (chroma), e.g.,
  \[ (H, 0 + i' \cdot C_{\max}, L_{\max} - i' \cdot (L_{\max} - L_{\min})). \]
This strategy is depicted in the left panel of Figure~\ref{fig:pal-s-construct} for a blue
hue $H = 260$ and different combinations of maximal chroma ($C_{\max} = 0$, $80$ and $100$,
respecitvely) and minimal luminance ($L_{\min} = 30$, $30$ and $50$, respectively).
The first two combinations are also shown in the first two rows of Figure~\ref{fig:pal-s-examples}.
The right panel of Figure~\ref{fig:pal-s-construct} shows that the exact same strategy
is not possible  for the green hue $H = 120$. While the gray colors without chroma can be chosen in 
the same way, there is a stronger trade-off between using dark colors (with low luminance)
and colorful colors (with high chroma). Hence, the second path from light gray to 
full green ends at a much lighter color with $L_{\min} = 75$.

@@

To increase the contrast between the colors in the palette even further, the ideas from
the previous sequential palettes can also be combined with qualitative palettes by
simultaneously varying the hue as well:
  \[ (H_2 - i \cdot (H_1 - H_2), C_{\max} - i' \cdot (C_{\max} - C_{\min}),
      L_{\max} - i'' \cdot (L_{\max} - L_{\min})). \]
One application is an HCL-based version of ``heat colors'' that increase from a light yellow
(e.g., $(90, 30, 90)$) to a full red (e.g., $(0, 100, 50)$). To make the change in hue
visible, the chroma needs to increase rather quickly for low values of $i$ and then only
slowly for higher values of $i$. This can be achieved by choosing an appropriate transformation
$i'$ for chroma and a different transformation $i''$ for the luminance.
Such a strategy is adopted for the palettes shown in the lower three rows in
Figure~\ref{fig:pal-s-examples} using different pairs of hues as well as different
chroma and luminance contrasts.

@@

Diverging palettes are also used for coding numerical information ranging in a certain
interval---however, this interval includes a neutral value. Examples for this include
residuals or correlations (both with the neutral value 0) or binary classification
probabilities (with neutral value 0.5) that could be visualized in mosaic plots
(see Figure~\ref{fig:haireye}), classification maps or model-based shading in maps
such as Figure~\ref{fig:nigeria}. Analogously to the previous
section, we suppose that we want to visualize an intensity or interestingness $i$
from the interval $[-1, 1]$ (without loss of generality).

@@

@@

Many statistical graphics---especially when displayed on a computer screen,
e.g., as in interactive usage, electronic papers or presentation slides---employ
colors to code information about a certain variable. Despite this
omnipresence of color, there is often only little guidance in statistical
software packages on how to choose a palette appropriate for a particular
visualization task---auspicious tools such as \pkg{ColorBrewer.org} notwithstanding.
We try to address this problem by suggesting color schemes for coding
categorical information (qualitative palettes) and numerical information
(sequential and diverging palettes) based on the perceptually-based
HCL color space.
